Testing Gravity using Type Ia Supernovae Discovered by LSST

### 2018-12-12 doc/article.tex
fix exponent in table header

\title{Testing Gravity using Type Ia Supernovae Discovered by LSST}
\author[0000-0001-6315-8743]{A.~G.~Kim}
\affiliation{    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{S.~BenZvi}
\affiliation{Department of Physics and Astronomy, University of Rochester, Rochester, NY 14627, USA}
\author{C.~Harper}
\affiliation{    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{C.~Ju}
\affiliation{    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{D.~Huterer}
\affiliation{Department of Physics, University of Michigan, 450 Church Street, Ann
Arbor, MI 48109, USA }

@@

\begin{abstract}
ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.2$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.  The increase in the number density of SNe~Ia, in parallel with improvements in the standardization of
their absolute magnitudes, now make them competitive probes of the growth of structure.  The peculiar velocity power spectrum derived from SNe~Ia
is sensitive to the product of the linear growth and amplitude of structure
$f\sigma_8$.  Cross-correlation with synergistic galaxy surveys further constrains $f \sigma_8$ and the galaxy bias.  Velocities
SNe~Ia in the next decade will measure the growth of structure  a powerful test of General Relativity in the local $z<0.2$ Universe.
\end{abstract}

@@

The growth of structure depends on the expansion history of the Universe, the nature and density of its contents, and gravity. 
It is therefore a powerful probe of cosmology and dark energy.  Growth of structure can be measured from the baryonic structures
in the Universe and from the peculiar velocities of test masses therein.
Peculiar velocities are the motions, on top of the cosmological expansion, caused by the gravitational attraction
and repulsion due to inhomogeneities in the Universe.  The peculiar velocity of an object wth a given absolute magnitude
is determined from its observer magnitude and redshift.  Given a background cosmology, the observed magnitude provides
an estimate of the cosmological redshift.  Peculiar velocity accounts for the difference between the cosmological and observer redshift.

Baryonic structures and peculiar velocities  provide a measurement of the combination $fD$.  $D$ is the ``linear growth factor'' that
gives the overall amplitude of  overdensities and the ``linear growth rate''
$$f \equiv \frac{d\ln{D}}{d\ln{a}}$$ is how that amplitude changes with redshift.  General Relativity predicts
$f \approx \Omega_M^\gamma$ (with $D$ determined accordingly) when $\gamma=0.55$, whereas other gravity models can be similarly described
with different values of $\gamma$
\citep{2007APh....28..481L}.  Thus growth of structure, through the measurement of $fD$, provides a test of General Relativity and breaks degeneracies
between gravitational and dark energy models that explain the accelerating expansion of the Universe.
The parameter $\sigma_8$, the  standard deviation of overdensities in 8$h^{-1}$Mpc spheres, is 
commonly used to normalize the
overall amplitude of  overdensities instead of $D$, and so the standard parameterization used by the community is $f\sigma_8$.

Baryonic structures are sensitive to $f\sigma_8$ through redshift space distortions (RSD), 
which gives the combination $(b + f \mu^2)\sigma_8$ where $b$ is the bias between the tracer and dark matter
and $\mu$ gives the angular separation.  Correlations between peculiar velocities are sensitive to  $f\sigma_8 (H_0 d_L)^{-1}$.
They are independent of the tracer bias because all sources of mass, predominantly dark matter, contribute to the dynamics. As
mentioned earlier, peculiar velocities are measured relative to the background cosmological expansion that
leads to the dependence on $H_0 d_L$.  The cross-correlation between structure and peculiar velocity within the same volume come from the same
overdensities and hence
provide independent information unsusceptible to sample variance \citep{2007PhRvL..99h1301G}.

Tracers with precisely known absolute magnitude provide precise cosmological redshifts and hence peculiar velocities.  
The current generation of peculiar velocity studies use $10^3-10^5$ galaxies with Fundamental Plane and Tully-Fisher distances \citep{2008AJ....135.1738M, 2014MNRAS.445.2677S,
2016AJ....152...50T}. 
Next generation surveys WALLABY \citep{2008ExA....22..151J} and TAIPAN \citep{2017PASA...34...47D} are designed to increase these sample size by an order of magnitude.
The galaxies have absolute magnitude uncertainties of $\sim 0.4$~mag.

@@

\begin{table}
   \centering
   %\topcaption{Table captions are better up top} % requires the topcapt package
   \begin{tabular}{|ccc|ccccc|} % Column formatting, @{} suppresses leading/trailing space
   \hline
$z_{max}$ & $\phi$ & $\sigma_{M}$ & $N_{gal}$ & $\bar{A}$ & $\sigma_A$ & $\bar{A} \sigma_A^{-1} N_{gal}^{0.5}$ & $\bar{A} \sigma_A^{-1} (18000/760)^{0.5}$ \\
\hline
0.07 & 0.65 & 0.08 & 392 &   1.34 &   0.49 &  0.137 &  13.21 \\
0.07 & 1.00 & 0.08 & 563 &   1.16 &   0.36 &  0.135 &  15.55 \\
0.07 & 1.00 & 0.10 & 563 &   1.27 &   0.48 &  0.112 &  12.88 \\
0.07 & 1.00 & 0.12 & 563 &   1.41 &   0.61 &  0.097 &  11.15 \\
0.07 & 1.00 & 0.15 & 563 &   1.64 &   0.82 &  0.084 &   9.74 \\
0.10 & 0.30 & 0.08 & 429 &   1.26 &   0.51 &  0.120 &  12.09 \\
0.10 & 0.50 & 0.08 & 683 &   1.02 &   0.42 &  0.093 &  11.84 \\
0.10 & 0.65 & 0.08 & 873 &   0.89 &   0.35 &  0.087 &  12.57 \\
0.10 & 0.65 & 0.10 & 873 &   1.01 &   0.46 &  0.075 &  10.72 \\
0.10 & 0.65 & 0.12 & 873 &   1.11 &   0.57 &  0.066 &   9.48 \\
0.10 & 1.00 & 0.08 & 1294 &   0.91 &   0.28 &  0.091 &  15.96 \\
0.10 & 1.00 & 0.10 & 1294 &   0.95 &   0.36 &  0.072 &  12.66 \\
0.10 & 1.00 & 0.12 & 1294 &   1.04 &   0.43 &  0.067 &  11.71 \\
0.15 & 0.30 & 0.08 & 1215 &   1.42 &   0.55 &  0.074 &  12.48 \\
0.15 & 0.30 & 0.12 & 1215 &   1.97 &   0.92 &  0.061 &  10.36 \\
0.15 & 0.50 & 0.08 & 1982 &   1.12 &   0.37 &  0.069 &  14.93 \\
0.15 & 0.65 & 0.08 & 2541 &   1.15 &   0.33 &  0.070 &  17.05 \\
0.15 & 0.65 & 0.10 & 2541 &   1.15 &   0.39 &  0.059 &  14.43 \\
0.15 & 0.65 & 0.12 & 2541 &   1.30 &   0.51 &  0.051 &  12.42 \\
0.15 & 1.00 & 0.08 & 3870 &   1.00 &   0.24 &  0.066 &  20.00 \\
0.20 & 0.20 & 0.08 & 2021 &   1.66 &   0.59 &  0.062 &  13.61 \\
0.20 & 0.30 & 0.08 & 3023 &   1.34 &   0.43 &  0.057 &  15.16 \\
0.20 & 0.30 & 0.10 & 3023 &   1.40 &   0.54 &  0.047 &  12.68 \\
0.20 & 0.30 & 0.12 & 3023 &   1.48 &   0.72 &  0.037 &   9.95 \\
0.20 & 0.50 & 0.08 & 5015 &   1.13 &   0.30 &  0.054 &  18.54 \\
0.20 & 0.65 & 0.08 & 6491 &   1.28 &   0.30 &  0.053 &  20.79 \\
0.20 & 0.65 & 0.10 & 6491 &   1.24 &   0.34 &  0.046 &  17.95 \\
0.20 & 0.65 & 0.12 & 6491 &   1.33 &   0.45 &  0.037 &  14.33 \\
0.20 & 1.00 & 0.08 & 9901 &   1.18 &   0.23 &  0.050 &  24.38 \\
0.20 & 1.00 & 0.10 & 9901 &   1.06 &   0.25 &  0.042 &  20.33 \\
0.20 & 1.00 & 0.12 & 9901 &   1.09 &   0.28 &  0.040 &  19.13 \\
    \hline
   \end{tabular}
   \caption{ston in $A$ or equivalently $f\sigma_8$ for surveys parameterized by the maximum redshift $z_{max}$,
   the fraction of total supernovae in 10 years ``fraction'', intrinsic magnitude dispersion $\sigma_{SN}$.  The number of galaxies
   $N_{gal}$, the signal $\bar{A}$ and its uncertainty $\sigma_A$ are for the cosmoDC2 (v1.0) 760 sq.~deg. volume only.
   The uncertainty ``per'' supernova is given by  $\sigma_A N_{gal}^{-0.5}$.  The ston, scaled to an LSST solid angle, is given as 
    $\bar{A} \sigma_A^{-1} (18000/760)^{0.5}$.
   \label{tab:subsets}}
\end{table}

>>>>

\title{Testing Gravity using Type Ia Supernovae Discovered by LSST}
\author[0000-0001-6315-8743]{A.~G.~Kim}
\affiliation{    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{S.~BenZvi}
\affiliation{Department of Physics and Astronomy, University of Rochester, Rochester, NY 14627, USA}
\author{C.~Harper}
\affiliation{    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}
\author{D.~Huterer}
\affiliation{Department of Physics, University of Michigan, 450 Church Street, Ann
Arbor, MI 48109, USA }
\author{C.~Ju}
\affiliation{    Physics Division, Lawrence Berkeley National Laboratory, 
    1 Cyclotron Road, Berkeley, CA, 94720}

@@

\begin{abstract}
ZTF today and LSST in the upcoming decade will increase the number of identified  $z<0.2$ Type~Ia supernovae (SNe~Ia)  from the hundreds to the
hundreds of thousands.  The increase in the number density of SNe~Ia, in parallel with improvements in the standardization of
their absolute magnitudes, now make them competitive probes of the growth of structure.  The peculiar velocity power spectrum
is sensitive to $f\sigma_8$,
the product of the linear growth and amplitude of density perturbations.  Cross-correlation with synergistic galaxy surveys further constrains $f \sigma_8$ and the galaxy bias. 
Thus, in the next decade the peculiar velocities of
SNe~Ia will provide the growth of structure  in the local $z<0.2$ Universe as a powerful test of General Relativity and other models of gravity.
\end{abstract}

@@

The growth of structure depends on the expansion history of the Universe, the nature and density of its contents, and gravity. 
It is therefore a powerful probe of cosmology and dark energy.  Growth of structure can be measured from the baryonic structures
in the Universe and from the peculiar velocities of test masses therein.
Peculiar velocities are the motions, on top of the cosmological expansion, caused by the gravitational attraction
and repulsion of densitiy inhomogeneities in the Universe.  The peculiar velocity of an object with a known absolute magnitude
is determined from its observer magnitude and redshift. For a given a background cosmology, the observed magnitude provides
an estimate of the cosmological redshift, the peculiar velocity is then the difference between the cosmological and observer redshifts.

Baryonic structures and peculiar velocities  provide a measurement of the combination $fD$.  $D$ is the ``linear growth factor'' that
gives the overall amplitude of  overdensities, and the ``linear growth rate''
$$f \equiv \frac{d\ln{D}}{d\ln{a}}$$ is how that amplitude changes with redshift.  General Relativity predicts
$f \approx \Omega_M^\gamma$ (with $D$ determined accordingly) with $\gamma=0.55$, whereas other gravity models can be similarly described
with different values of $\gamma$
\citep{2007APh....28..481L}.  The growth of structure, through the measurement of $fD$, provides a test of General Relativity and breaks degeneracies
between gravitational and dark energy models that explain the accelerating expansion of the Universe.
The parameter $\sigma_8$, the  standard deviation of overdensities in 8$h^{-1}$Mpc spheres, is 
commonly used in place of $D$ to normalize the
overall amplitude of  overdensities, so the standard parameterization used by the community is $f\sigma_8$.

Baryonic structures are sensitive to $f\sigma_8$ through redshift space distortions (RSD), 
which gives the combination $(b + f \mu^2)\sigma_8$ where $b$ is the bias between the tracer and dark matter
and $\mu$ gives the angular separation.  Correlations between peculiar velocities are sensitive to  $f\sigma_8 (H_0 d_L)^{-1}$.
Although the velocities may be measured off of biased tracers, the tracers' dynamics are driven by all mass (including dark matter) and are so bias-free.
As
mentioned earlier, peculiar velocities are measured relative to the background cosmological expansion that
leads to the dependence on $H_0 d_L$. 
Baryonic structures and peculiar velocities within the same volume are induced by the same overdensities  making their cross-correlation 
insensitive to sample variance \citep{2007PhRvL..99h1301G}: galaxy and peculiar velocity surveys provide synergistic constraints more powerful
than their naive sum.

The probative power of a specific peculiar-velocity tracer primarily depends on its number density and the precision to which its absolute magnitudes are known.
The current generation of peculiar velocity studies use $10^3-10^5$ galaxies with Fundamental Plane and Tully-Fisher distances \citep{2008AJ....135.1738M, 2014MNRAS.445.2677S,
2016AJ....152...50T}. 
Next generation surveys WALLABY \citep{2008ExA....22..151J} and TAIPAN \citep{2017PASA...34...47D} are designed to increase these sample size by an order of magnitude.
These galaxies have absolute magnitude uncertainties of $\sim 0.4$~mag.

@@

\begin{table}
   \centering
   %\topcaption{Table captions are better up top} % requires the topcapt package
   \begin{tabular}{|ccc|ccccc|} % Column formatting, @{} suppresses leading/trailing space
   \hline
$z_{max}$ & $\phi$ & $\sigma_{M}$ & $N_{gal}$ & $\bar{A}$ & $\sigma_A$ & $\bar{A} \sigma_A^{-1} N_{gal}^{-0.5}$ & $\bar{A} \sigma_A^{-1} (18000/760)^{0.5}$ \\
\hline
0.07 & 0.65 & 0.08 & 392 &   1.34 &   0.49 &  0.137 &  13.21 \\
0.07 & 1.00 & 0.08 & 563 &   1.16 &   0.36 &  0.135 &  15.55 \\
0.07 & 1.00 & 0.10 & 563 &   1.27 &   0.48 &  0.112 &  12.88 \\
0.07 & 1.00 & 0.12 & 563 &   1.41 &   0.61 &  0.097 &  11.15 \\
0.07 & 1.00 & 0.15 & 563 &   1.64 &   0.82 &  0.084 &   9.74 \\
0.10 & 0.30 & 0.08 & 429 &   1.26 &   0.51 &  0.120 &  12.09 \\
0.10 & 0.50 & 0.08 & 683 &   1.02 &   0.42 &  0.093 &  11.84 \\
0.10 & 0.65 & 0.08 & 873 &   0.89 &   0.35 &  0.087 &  12.57 \\
0.10 & 0.65 & 0.10 & 873 &   1.01 &   0.46 &  0.075 &  10.72 \\
0.10 & 0.65 & 0.12 & 873 &   1.11 &   0.57 &  0.066 &   9.48 \\
0.10 & 1.00 & 0.08 & 1294 &   0.91 &   0.28 &  0.091 &  15.96 \\
0.10 & 1.00 & 0.10 & 1294 &   0.95 &   0.36 &  0.072 &  12.66 \\
0.10 & 1.00 & 0.12 & 1294 &   1.04 &   0.43 &  0.067 &  11.71 \\
0.15 & 0.30 & 0.08 & 1215 &   1.42 &   0.55 &  0.074 &  12.48 \\
0.15 & 0.30 & 0.12 & 1215 &   1.97 &   0.92 &  0.061 &  10.36 \\
0.15 & 0.50 & 0.08 & 1982 &   1.12 &   0.37 &  0.069 &  14.93 \\
0.15 & 0.65 & 0.08 & 2541 &   1.15 &   0.33 &  0.070 &  17.05 \\
0.15 & 0.65 & 0.10 & 2541 &   1.15 &   0.39 &  0.059 &  14.43 \\
0.15 & 0.65 & 0.12 & 2541 &   1.30 &   0.51 &  0.051 &  12.42 \\
0.15 & 1.00 & 0.08 & 3870 &   1.00 &   0.24 &  0.066 &  20.00 \\
0.20 & 0.20 & 0.08 & 2021 &   1.66 &   0.59 &  0.062 &  13.61 \\
0.20 & 0.30 & 0.08 & 3023 &   1.34 &   0.43 &  0.057 &  15.16 \\
0.20 & 0.30 & 0.10 & 3023 &   1.40 &   0.54 &  0.047 &  12.68 \\
0.20 & 0.30 & 0.12 & 3023 &   1.48 &   0.72 &  0.037 &   9.95 \\
0.20 & 0.50 & 0.08 & 5015 &   1.13 &   0.30 &  0.054 &  18.54 \\
0.20 & 0.65 & 0.08 & 6491 &   1.28 &   0.30 &  0.053 &  20.79 \\
0.20 & 0.65 & 0.10 & 6491 &   1.24 &   0.34 &  0.046 &  17.95 \\
0.20 & 0.65 & 0.12 & 6491 &   1.33 &   0.45 &  0.037 &  14.33 \\
0.20 & 1.00 & 0.08 & 9901 &   1.18 &   0.23 &  0.050 &  24.38 \\
0.20 & 1.00 & 0.10 & 9901 &   1.06 &   0.25 &  0.042 &  20.33 \\
0.20 & 1.00 & 0.12 & 9901 &   1.09 &   0.28 &  0.040 &  19.13 \\
    \hline
   \end{tabular}
   \caption{ston in $A$ or equivalently $f\sigma_8$ for surveys parameterized by the maximum redshift $z_{max}$,
   the fraction of total supernovae in 10 years ``fraction'', intrinsic magnitude dispersion $\sigma_{SN}$.  The number of galaxies
   $N_{gal}$, the signal $\bar{A}$ and its uncertainty $\sigma_A$ are for the cosmoDC2 (v1.0) 760 sq.~deg. volume only.
   The uncertainty ``per'' supernova is given by  $\sigma_A N_{gal}^{-0.5}$.  The ston, scaled to an LSST solid angle, is given as 
    $\bar{A} \sigma_A^{-1} (18000/760)^{0.5}$.
   \label{tab:subsets}}
\end{table}
